Kindly Bent to Free Us

### 2019-11-18 kindly/infer/extra.tex
Move explanation of Let in the appendix.

In this appendix, we provide the complete type inference rules
and show that our type inference algorithm is sound and complete.
The constraints rules are already shown in \cref{inference}.
Kind inference is presented in \cref{rules:kinding:full}.
The type inference rules are shown in \cref{rules:typing:full}.
The various theorems and their proofs are direct adaptations
of the equivalent statements for HM(X) \citep{sulzmann1997proofs}.

@@

>>>>

In this appendix, we provide the complete type inference rules
and show that our type inference algorithm is sound and complete.
The constraints rules are already shown in \cref{inference}.
Kind inference is presented in \cref{rules:kinding:full}
and the detailed treatment of let-bindings in
\cref{infer:let}.
The type inference rules are shown in \cref{rules:typing:full}.
The various theorems and their proofs are direct adaptations
of the equivalent statements for HM(X) \citep{sulzmann1997proofs}.

@@

\subsection{Generalization and constraints}
\label{infer:let}

% So far, constraints have only been composed of a list of kind inequalities.
% The internal language of constraints used for inference is in fact
% richer.
% To illustrate this, we show the inference rule for let-bindings
% in \cref{rule:infer:let}.
The {\sc Let} rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
the right environment, since it is not available outside of the expression
$e_2$, and should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
for the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
$\Cproj{x}{D}$ where $x$ can be either a type or a kind variable, which
allow the creation of local variables inside the constraints.
This allows us to encapsulate all the quantified variables in the global constraints.
It also reflects the fact that there
should exist at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

### 2019-11-18 kindly/infer/infer.tex
Move explanation of Let in the appendix.

\subsubsection{Usage environments}

@@

\subsubsection{Constraint normalization}

@@

\subsubsection{Environments and bindings}
\label{infer:envs}
%
\begin{figure*}[tb]
  \begin{mathpar}
    \ruleIVar
    \and
    \ruleIAbs
    \and
    \ruleILet
    \hfill
    \ruleIRegion
    \and
    \ruleIPair
    \and
    \inferrule{}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \text{if } (x\in\Sv) \text{ then} \Ctrue \text{else }
      \Cleq{\sigma}{\kaff_\infty}
      % \begin{dcases}
      %   \Ctrue&\text{if } x\in\Sv\\
      %   \Cleq{\sigma}{\kaff_\infty}&\\
      % \end{dcases}
      \\\\\\
      \generalize{C}{\E}{\tau} =
      (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
      \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
      \text{ where}\\\\
      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \vspace{-15pt}
  \caption{Selected inference rules}
  \label{rule:infer:envs}
  \label{rule:infer:envrules}
  \label{rule:infer:let}
\end{figure*}
%
In our syntax-directed system, we ensured
that linear variables are not discarded at the \emph{leaves}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leaves, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax-directed type system,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which is false only if weakening
is necessary but impossible due to linearity.
The $\Weaken$ constraint is introduced at each binding point in the program
such as let-bindings and pattern matching.

@@

We now present the inference version
of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
While the rules are quite similar to the original version, the usage
environment $\Sv$ is now an \emph{output} of the algorithm.
As such, we use the ``inference'' version of the relations on
the environment,
$\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$
and $\bregion{C}{\inP{x}}{\Sv}{\inP\Sv'}$,
which returns the necessary constraints.
As appropriate for inference, these relations
are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
We then collect all constraints and normalize them.

\subsubsection{Generalization and constraints}

So far, constraints have only been composed of a list of kind inequalities.
The internal language of constraints used for inference is in fact
richer.
To illustrate this, we show the inference rule for let-bindings
in \cref{rule:infer:let}.
This rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
the right environment, since it is not available outside of the expression
$e_2$, and should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
for the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
$\Cproj{x}{D}$ where $x$ can be either a type or a kind variable, which
allow the creation of local variables inside the constraints.
This allows us to encapsulate all the quantified variables in the global constraints.
It also reflects the fact that there
should exist at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

>>>>

\paragraph{Usage environments}

@@

\paragraph{Constraint normalization}

@@

\subsubsection{Environments and bindings}
\label{infer:envs}
%
\begin{figure*}[tb]
  \begin{mathpar}
    \ruleIVar
    \and
    \ruleIAbs
    \and
    % \ruleILet
    % \hfill
    \ruleIRegion
    \and
    \ruleIPair
    \and
    \inferrule{}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \text{if } (x\in\Sv) \text{ then} \Ctrue \text{else }
      \Cleq{\sigma}{\kaff_\infty}
      % \begin{dcases}
      %   \Ctrue&\text{if } x\in\Sv\\
      %   \Cleq{\sigma}{\kaff_\infty}&\\
      % \end{dcases}
      % \\\\\\
      % \generalize{C}{\E}{\tau} =
      % (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
      % \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
      % \text{ where}\\\\
      % \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \vspace{-15pt}
  \caption{Selected inference rules -- $\inferW{\Sv}{(C,\unif)}{\inP{\E}}{\inP{e}}{\tau}$}
  \label{rule:infer:envs}
  \label{rule:infer:envrules}
  \label{rule:infer:let}
\end{figure*}
%
In our syntax-directed system, we ensured
that linear variables are not discarded at the \emph{leaves}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leaves, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax-directed type system,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which is false only if weakening
is necessary but impossible due to linearity.
The $\Weaken$ constraint is introduced at each binding point in the program
such as let-bindings and pattern matching.

@@

The inference version
of the {\sc Pair} and {\sc Region} rules is shown in \cref{rule:infer:envrules}.
While the rules are quite similar to the original version, the usage
environment $\Sv$ is now an \emph{output} of the algorithm.
As such, we use the ``inference'' version of the relations on
the environment,
$\bsplit{C}{\Sv}{\inP{\Sv_1}}{\inP{\Sv_2}}$
and $\bregion{C}{\inP{x}}{\Sv}{\inP\Sv'}$,
which returns the necessary constraints.
As appropriate for inference, these relations
are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
We then collect all constraints and normalize them.
